 

### 2021-07-12 vertical-thermal-review.Rmd
a few changes
... nothing very productive...

Generally, the pattern of fundamental trait adaptation to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. 
However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions. Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].

@@

@@

**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality. There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors.Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a]. However, growth also declines as trees age. This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts]. Vertical biophysical conditions are bound to influence growth and whole tree functioning.

>>>>

Generally, the pattern of fundamental trait adaptation to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. 
However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions. 
8Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].*

@@

Numerous traits vary between understory and canopy species. (*this needs to be reconciled with interspecific portion of traits section*)

@@

**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* 
Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors. 
Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a]. 
However, growth also declines as trees age. 
This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts]. 
Vertical biophysical conditions are bound to influence growth and whole tree functioning.
